TacticToe: Learning the Game of Tactics

### 2018-01-31 tactictoe_jv.tex
recording: wip

Because of the large number of \sml constructions, we only describe the effect 
of these steps in a running example that is representative of a typical proof 
script. And we comment on the main steps that transforms the original proof 
script into a modified version that is able to record the code of a tactic and 
the goal(s) it processed during the proof. We omit more involved constructions 
such as the globalization of infix operators.

@@

@@

\begin{example}\label{ex:global} (Globalization)
\begin{lstlisting}[language=SMLSmall]
Tactic.STRIP_TAC THEN
Prim_rec.INDUCT_THEN (DB.fetch "list" "list_INDUCT") Tactic.ASSUME_TAC THEN
Rewrite.ASM_REWRITE_TAC [DB.fetch "list" "MAP", DB.fetch "list" "APPEND"]
\end{lstlisting}
\end{example}

@@

@@

The function save(stac,goal) save the pair to disk increasing the number of 
tactic-goal pairs in our database.

%\subsection{Data Transfer}
%We describe here the format preferred to save and load the 
%data recorded by the function \texttt{save}. 
%During saving, we add fields to the input pair $(tactic,goal)$ to record 
%precomputed information in order to make the prediction faster. The final 
%feature vector is the tuple $(tactic,time,goal,goal\_list,features)$.
%The time $time$ taken by $tactic$ is not currently used by our 
%predictor. The list of goals $goal list$ is the output of $tactic$ on 
%$goal$. And the features $features$ of the goal $goal$ are extracted and 
%hashed 
%as described in Section~\ref{sec:features}. 
%
%\begin{remark}
%To simplify the description of our machine learning algorithms in future 
%sections, we consider a tactic feature vector to be a pair $(tactic,goal)$ and 
%only 
%explicitly rely on the 
%other fields when necessary.
%\end{remark}
%
%We now discuss the type of file we want to store our feature vectors on and
%how we plan to write and read \sml values. We start by taking 
%inspiration from an existing implementation in \holfour.
%
%To avoid the computational cost of re-proving every theorems at the begining 
%of 
%each session, theorem values in \holfour are stored.
%They are encoded them in to a string by calling \texttt{Term.write\_raw}.
%Conversely, the function \texttt{Term.read\_raw} transforms the string back 
%into the original term.
%To have theorems binded to sml identifiers, it is necessary to create a 
%\sml file where those binders are declared. And for simplicity sake, all 
%theory 
%data was fed into this \sml file. However, it is
%more efficient to store theory data into a text file and make a custom 
%parser which can be run faster than the \polyml compiler.
%Thanks to this insight, we have 
%modified the way in which \holfour theories are saved to include only \sml 
%bindings in a \sml file. All other theory data is now stored separately 
%into a text file. After integrating the changes, \holfour theories load three 
%times faster in average.
%
%Our \tactictoe data management system do not need to create new \sml
%bindings for the user. So we can store all \sml values into a text file which 
%can be loaded rapidly.
%Storing time, tactic string and features is easy since they are respectively 
%floats, strings and integers. And because a goal is composed of terms, we can 
%re-use the \holfour theory data reading and writing functions 
%for the two other fields.
%
%Finally, we mention some limitations that comes with 
%the need to record tactics as a string compared to being able to store a \sml 
%value. Access to the name of the tactic enhances machine 
%learning ability and user interaction. Yet, because some tactics 
%depends on the state of \holfour, for example the state of the global 
%simpset, a tactic is re-parsed before each proof attempt to be up-to-date.
%As a result, a tactic may have different effect on goals than suggested by the 
%database which could compromise the prediction accuracy. In practice, this 
%should rarely happen. Indeed, most tactics are stateless and an update on a 
%stateful tactic is usually improving it.

@@

\todoi{Motivation: + minimal effort to integrate the tactictoe results in proof 
scripts.}

When \tactictoe finds a proof of a goal $g_{root}$, it returns a search tree 
$\mathfrak{T}$ where the root node containing $g_{root}$ is solved.
In order to transform this search tree into a proof script, we need to extract 
the tactics that contributed to the proof and combine them using tactic 
combinators.
\todoi{Define tactic combinators?}
By the design of the search, a single tactic combinator, \texttt{THENL}, is 
sufficient. It combines a tactic $t$ with a list of subsequent ones, in such a 
way that after $t$ is called, for each created goal a respective 
tactic from the list is called.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by subtituting \texttt{THENL} by \texttt{THEN} when the 
list of goals is a singleton and removing \texttt{THENL []} during the 
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improve the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.
\todoi{Not sure if this reference is appropriate anymore}

>>>>

Because of the large number of \sml constructions, we only describe the effect 
of these steps on a running example that contains selected parts of the theory
of lists.

%We comment on the treatment of global and local identifiers but  omit more 
%involved constructions 
%such as the globalization of infix operators.

@@

The first line of this theory open modules (called structures in \sml). Each of 
them contains a list of global identifiers which become directly accessible in 
the rest of the theory.
A local identifier \texttt{LIST\_INDUCT\_TAC} is declared after, which is a 
tactic that performs induction on lists. And even further, the theorem 
\texttt{MAP\_APPEND} is proven.
The global tactic \texttt{STRIP\_TAC} first removes universal quantifiers. Then,
the goal is split into a base case and a inductive case. And finally, both of 
theses cases are solved by \texttt{ASM\_REWRITE\_TAC [MAP, APPEND]}, which 
rewrites assumptions with the help of theorems previously declared in this 
theory.

We first parse the theory to extract proof scripts. Each of them is found in 
the third argument of a call to \texttt{store\_thm}. The result of proof script 
extraction for the running example in presented in Example~\ref{ex:r0}.

@@

In the next phase, we globalize identifiers of the proof scripts. 
Infix operators such as \texttt{THEN} need to be processed in a special way so 
that they can keep their infixity after globalization. For simplicity, the 
globalization of infix operators is omitted in Example~\ref{ex:r1}.
In this example, the three main cases that can happen during the globalization
are depicted. The first one is the globalization of identifiers declared in 
modules. The global identifiers \texttt{STRIP\_TAC} and 
\texttt{ASM\_REWRITE\_TAC} are prefixed by their module \texttt{Tactic}. In 
this way, they will still be interpretable whether \texttt{Tactic} was open or 
not. The local identifier \texttt{LIST\_INDUCT\_TAC} is replaced by its 
definition which happens to contain two global identifiers.
The previous paragraph describe the globalization for all identifiers except 
local theorems.  We do not replace a local
theorem by its definition because we want to avoid unfolding proof scripts into 
other proof scripts.
If the theorem is stored in the database of the current theory, we 
can obtain the theorem value by calling \texttt{DB.fetch}. This happens for the 
three theorems in our script. The internal database of \holfour is available 
across \holfour developments. 
Otherwise the globalization process fails and the local theorem identifier is 
kept unchanged. A recorded tactic with an unchanged local theorem as argument 
is only interpretable inside the current theory.

\begin{example}\label{ex:r1} (Globalization)
\begin{lstlisting}[language=SMLSmall]
Tactic.STRIP_TAC THEN
Prim_rec.INDUCT_THEN (DB.fetch "list" "list_INDUCT") Tactic.ASSUME_TAC THEN
Rewrite.ASM_REWRITE_TAC [DB.fetch "list" "MAP", DB.fetch "list" "APPEND"]
\end{lstlisting}
\end{example}

\todo{rewrite this paragraph}

@@

\todo{rewrite this paragraph}

@@

The function \texttt{save} writes the tactic-goal pairs to disk increasing the 
number of entries in our database of tactics. The function 
\texttt{tactic\_of\_sml} interprets \sml code \texttt{stac}. The tactic is then 
applied to the input goal which replicates the original behavior.

@@

When \tactictoe finds a proof of a goal $g_{root}$, it returns a search tree 
$\mathfrak{T}$ where the root node containing $g_{root}$ is solved.
In order to transform this search tree into a proof script, we need to extract 
the tactics that contributed to the proof and combine them using tactic 
combinators.

By the design of the search, a single tactic combinator, \texttt{THENL}, is 
sufficient. It combines a tactic $t$ with a list of subsequent ones, in such a 
way that after $t$ is called, for each created goal a respective 
tactic from the list is called.

@@

The extracted proof script of $\mathfrak{T}$ is $P_{goal}(g_{root})$.
We minimally improve it by subtituting \texttt{THENL} by \texttt{THEN} when the 
list of goals is a singleton and removing \texttt{THENL []} during the 
extraction phase.
Further post-processing such as
eliminating unnecessary tactics and theorems has been developed and
improve the user experience greatly~\cite{DBLP:conf/sefm/Adams15}.
